### 2015-06-06 chapters/ipa_meets_unicode.tex
Updated Pitfall IPA homoglyphs

As explained in Section~\ref{pitfall-blocks}, the Unicode Standard code space is
subdivided into character blocks, which generally encode characters from a
single script. However, as is illustrated by the IPA, characters that form a
single writing system may be dispersed across several different character
blocks. With its diverse collection of symbols from various scripts and
diacritics, the IPA is spread across 13 blocks in the Unicode
Standard:\footnote{This number of blocks depends on whether only IPA-sanctioned
symbols are counted or if the phonetic symbols commonly found in the literature
are also included, see \cite[Appendix C]{Moran2012}}
\begin{itemize}
	\item Basic Latin (30 characters), e.g. <a b c d e \ldots> 
	\item Latin-1 Supplement (4 characters): <æ ç ð ø> 
	\item Latin Extended-A (3 characters): <ħ ŋ œ> 
	\item Latin Extended-B (5 characters): <ǀ ǁ ǂ ǃ ȵ> 
	\item IPA Extensions (70 characters), e.g. <ɐ ɑ ɔ \ldots> 
	\item Spacing Modifier Letters (20 characters), e.g. <ʰ ʷ ˥ \ldots> 
	\item Combining Diacritical Marks (33 characters), e.g. <{\large \ \ ̝\ \ ̥\ \ ̪ }\ldots> 
	\item Greek and Coptic (3 characters): <β θ χ> 
	\item Phonetic Extensions (2 characters): <{\small \fontspec{CharisSIL}ᴅ ᴴ}> 
	\item Phonetic Extensions Supplement (3 characters): <{\small \fontspec{CharisSIL}ᶑ ᶾ ᶣ}> 
	\item Superscripts and Subscripts (1 character): <ⁿ> 
	\item Arrows (4 characters): <↑ ↓ ↗ ↘>
	\item Latin Extended-C (1 character): <{\small \fontspec{CharisSIL}ⱱ}> 
\end{itemize}

@@

Another problem is the large number of homoglyphs, i.e.~different characters
that have highly similar glyphs (or even completely identical, depending on the
font rendering). For example, a speaker of Russian should ideally not use the
\textsc{cyrillic small letter a} at code point \uni{0430} for IPA
transcriptions, but instead the \textsc{latin small letter a} at code point
\uni{0061}, although visually they are mostly indistinguishable, and the
Cyrillic character is more easily typed on a Cyrillic keyboard. Another example
we commonly encounter is the use of \textsc{latin small letter g} at \uni{0067},
instead of the sanctioned Unicode Standard IPA character for the voiced velar
stop \textsc{latin small letter script g} at \uni{0261}. One begins to question
whether this issue is at all apparent to the working linguist, or if they simply
use the \uni{0067} because it is easily keyboarded and thus saves time, whereas
the latter must be cumbersomely inserted as a special symbol in most
software.\footnote{This issue was recently addressed by the International
Phonetic Association, which has taken the stance that both the keyboard
\textsc{latin small letter g} and the \textsc{latin small letter script g} are
valid input characters for the voiced velar plosive. Unfortunately, this
decision further introduces ambiguity for linguists trying to adhere to a strict
Unicode Standard IPA encoding.}

Furthermore, on the one hand even linguists are unlikely to distinguish between
the <ə> \textsc{latin small letter schwa} at code point \uni{0259} and <ǝ>
\textsc{latin small letter turned e} at \uni{01DD}. On the other hand,
non-linguists are unlikely to distinguish any semantic difference between an
open back unrounded vowel <ɑ>, the \textsc{latin small letter alpha} at
\uni{0251}, and the open front unrounded vowel <a>, \textsc{latin small letter
a} at \uni{0061}. But even among linguists this distinction leads to problems.
For example, as pointed out by \citet{Mielke2009}, there is a problem stemming
from the fact that about 75\% of languages are reported to have a five-vowel
system \citep{Maddieson1984}. Historically, linguistic descriptions tend not to
include precise audio recording and measurements of formants, so this may lead
one to ask if the many characters that are used in phonological description
reflects a \textsc{transcriptional bias}. The common use of in transcriptions
could be in part due to the ease of typing the letter on an English keyboard (or
for older descriptions, the typewriter). We found it to be exceedingly rare that
a linguist uses for a low back unrounded vowel.\footnote{One example is
\citet[75]{Vidal2001a}, in which the author states: ``The definition of Pilagá
/a/ as {[}+back{]} results from its behavior in certain phonological contexts.
For instance, uvular and pharyngeal consonants only occur around /a/ and /o/.
Hence, the characterization of /a/ and /o/ as a natural class of (i.e.,
{[}+back{]} vowels), as opposed to /i/ and /e/.''} They simply use $<$a$>$ as
long as there is no opposition to $<$ɑ$>$.\footnote{See Thomason's Language Log
post, ``Why I don't love the International Phonetic Alphabet'', at:
\url{http://itre.cis.upenn.edu/\textasciitilde{}myl/languagelog/archives/005287.html}.}
Making things even more problematic, there is an old typographic tradition that
the double-story uses a single-story in italics. This leads to the unfortunate
effect that in most well-designed fonts the italics of and use the same glyph.
If this distinction has to be kept upright in italics, the only solution we can
currently offer is to use \textit{slanted} glyphs (i.e.~artificially italicized
glyphs) instead of real italics (i.e.~special italics glyphs designed by a
typographer).\footnote{For example, the widely used IPA font Doulos SIL
(\url{http://scripts.sil.org/cms/scripts/page.php?item\_id=DoulosSIL}) does not
have real italics. This leads some word-processing software, like Microsoft
Word, to produce slanted glyphs instead. That particular combination of font and
software application will thus lead to the desired effect. However, note that
when the text is transferred to another font (i.e.~one that includes real
italics) and/or to another software application (like Apple Pages, which does
not perform slanting), then this visual appearance will be lost. In this case we
are thus still in the pre-Unicode situation in which the choice of font and
rendering software actually matters. The ideal solution from a linguistic point
of view would be the introduction of a new IPA code point for a different kind
of which explicitly specifies that it should still be rendered as a double-story
character when italicized. After informal discussion with various Unicode
players, our impression is that this highly restricted problem is not
sufficiently urgent to introduce even more <a>-like characters in Unicode (which
already lead to much confusion, see Section \ref{pitfall-homoglyphs}). This is a
clear situation in which the Unicode Consortium is not just thinking about
linguists, but has a more wide-ranging practical view to consider.}

Some other homoglyphs related to encoding IPA in the Unicode Standard are:
\begin{itemize}
	\item The uses of the apostrophe has led to long discussions on the Unicode
       Standard email list. An English keyboard inputs
       \textless{}`\textgreater{} APOSTROPHE at \uni{0027}, although the
       \textit{preferred} Unicode apostrophe is the \textless{}'\textgreater{}
       \textsc{right single quotation mark} at \uni{2019}. Yet the glottal
       stop/glottalization/ejective marker is another completely different
       character, the \textsc{modifier letter apostrophe} at \uni{02BC}, but
       unfortunately looks mostly highly similar to \uni{2019}. 
	\item There is ambiguous encoding of IPA segments within the Unicode Standard.
       An example is the \uni{02C1} \textsc{modifier letter reversed glottal
       stop} vs the \uni{02E4} \textsc{modifier letter small reversed glottal
       stop} . Both are denoted in the Unicode Standard as the
       \textsc{pharyngealized diacritic} and both appear in various resources
       representing phonetic data online. This is thus an example for which the
       Unicode Standard does not solve the linguistic standardization problem. 
	\item There is at least one case in which the character name assigned by the
       Unicode Consortium does not match the IPA's description: in the Unicode
       Standard at \uni{01C3} is labeled \textsc{latin letter retroflex click},
       but in IPA is an alveolar or postalveolar click (not retroflex). This
       naming is probably best seen as a simple error in the Unicode Standard.
\end{itemize}

@@

It is not just the Unicode Standard that offers multiple options for encoding 
the IPA. Even the IPA itself is not consistent in how specific meaningful 
transcriptions have to be encoded. There are a few cases in which the IPA 
explicitly allows for different options of transcribing the same phonetic 
content
\begin{itemize}
  \item tone marks (prefer tone symbols)
  \item marking of <g>
  \item marking of voiceless with ring diacritic
  \item tie bar
\end{itemize}

>>>>

As explained in Section~\ref{pitfall-blocks}, the Unicode Standard code space is
subdivided into character blocks, which generally encode characters from a
single script. However, as is illustrated by the IPA, characters that form a
single writing system may be dispersed across several different character
blocks. With its diverse collection of symbols from various scripts and
diacritics, the IPA is spread across 13 blocks in the Unicode
Standard:\footnote{This number of blocks depends on whether only IPA-sanctioned
symbols are counted or if the phonetic symbols commonly found in the literature
are also included, see \cite[Appendix C]{Moran2012}.}

\begin{itemize}
	\item Basic Latin (30 characters), e.g. <a b c d e> 
	\item Latin-1 Supplement (4 characters): <æ ç ð ø> 
	\item Latin Extended-A (3 characters): <ħ ŋ œ> 
	\item Latin Extended-B (5 characters): <ǀ ǁ ǂ ǃ ȵ> 
	\item IPA Extensions (70 characters), e.g. <ɐ ɑ ɔ> 
	\item Spacing Modifier Letters (20 characters), e.g. <ʰ ʷ ˥> 
	\item Combining Diacritical Marks (33 characters), e.g. <{\large \ \ ̝\ \ ̥\ \ ̪ }> 
	\item Greek and Coptic (3 characters): <β θ χ> 
	\item Phonetic Extensions (2 characters): <{\small \fontspec{CharisSIL}ᴅ ᴴ}> 
	\item Phonetic Extensions Supplement (3 characters): <{\small \fontspec{CharisSIL}ᶑ ᶾ ᶣ}> 
	\item Superscripts and Subscripts (1 character): <ⁿ> 
	\item Arrows (4 characters): <↑ ↓ ↗ ↘>
	\item Latin Extended-C (1 character): <{\small \fontspec{CharisSIL}ⱱ}> 
\end{itemize}

@@

Another problem is the large number of homoglyphs, i.e.~different characters
that have highly similar glyphs (or even completely identical, depending on the
font rendering). For example, a speaker of Russian should ideally not use the
<а> \textsc{cyrillic small letter a} at code point \uni{0430} for IPA
transcriptions, but instead use the <a> \textsc{latin small letter a} at code point
\uni{0061}, although visually they are mostly indistinguishable, and the
Cyrillic character is more easily typed on a Cyrillic keyboard.

Furthermore, even linguists are unlikely to distinguish between
the <ə> \textsc{latin small letter schwa} at code point \uni{0259} and <ǝ>
\textsc{latin small letter turned e} at \uni{01DD}. Conversely,
non-linguists are unlikely to distinguish any semantic difference between an
open back unrounded vowel <ɑ> \textsc{latin small letter alpha} at
\uni{0251}, and the open front unrounded vowel <a> \textsc{latin small letter
a} at \uni{0061}. But even among linguists this distinction leads to problems.
For example, as pointed out by \citet{Mielke2009}, there is a problem stemming
from the fact that about 75\% of languages are reported to have a five-vowel
system \citep{Maddieson1984}. Historically, linguistic descriptions tend not to
include precise audio recording and measurements of formants, so this may lead
one to ask if the many characters that are used in phonological description
reflects a transcriptional bias. The common use of <a> in transcriptions
could be in part due to the ease of typing the letter on an English keyboard (or
for older descriptions, the typewriter). We found it to be exceedingly rare that
a linguist uses <ɑ> for a low back unrounded vowel.\footnote{One example is
\citet[75]{Vidal2001a}, in which the author states: ``The definition of Pilagá
/a/ as {[}+back{]} results from its behavior in certain phonological contexts.
For instance, uvular and pharyngeal consonants only occur around /a/ and /o/.
Hence, the characterization of /a/ and /o/ as a natural class of (i.e.,
{[}+back{]} vowels), as opposed to /i/ and /e/.''} They simply use <a> as
long as there is no opposition to <ɑ>.\footnote{See Thomason's Language Log
post, ``Why I don't love the International Phonetic Alphabet'' at:
\url{http://itre.cis.upenn.edu/~myl/languagelog/archives/005287.html}.}

Making things even more problematic, there is an old typographic tradition that
the double-story <a> uses a single-story <ɑ> in italics. This leads to the
unfortunate effect that in most well-designed fonts the italics of <a> and <ɑ>
use the same glyph. If this distinction has to be kept upright in italics, the
only solution we can currently offer is to use \textsc{slanted} glyphs
(i.e.~artificially italicized glyphs) instead of real italics (i.e.~special
italics glyphs designed by a typographer).\footnote{For example, the widely used
IPA font Doulos SIL
(\url{http://scripts.sil.org/cms/scripts/page.php?item\_id=DoulosSIL}) does not
have real italics. This leads some word-processing software, like Microsoft
Word, to produce slanted glyphs instead. That particular combination of font and
software application will thus lead to the desired effect distinguishing <a>
from <ɑ>. However, note that when the text is transferred to another font
(i.e.~one that includes real italics) and/or to another software application
(like Apple Pages, which does not perform slanting), then this visual appearance
will be lost. In this case we are thus still in the pre-Unicode situation in
which the choice of font and rendering software actually matters. The ideal
solution from a linguistic point of view would be the introduction of a new IPA
code point for a different kind of which explicitly specifies that it should
still be rendered as a double-story character when italicized. After informal
discussion with various Unicode players, our impression is that this highly
restricted problem is not sufficiently urgent to introduce even more <a>-like
characters in Unicode (which already lead to much confusion, see Section~
\ref{pitfall-homoglyphs}).}

Some other homoglyphs related to encoding IPA in the Unicode Standard are:

\begin{itemize}
	\item The uses of the apostrophe has led to long discussions on the Unicode
       Standard email list. An English keyboard inputs <{\fontspec{Monaco}'}> \textsc{apostrophe}
       at \uni{0027}, although the preferred Unicode apostrophe is the <\ ' >
       \textsc{right single quotation mark} at \uni{2019}.
     \item The glottal stop/glottalization/ejective marker is another completely
        different character <{\large ʼ}>, the \textsc{modifier letter apostrophe} at
        \uni{02BC}, but unfortunately looks mostly highly similar to
        \uni{2019}. 
	\item Another problem is the <ˁ> \textsc{modifier letter reversed
       glottal stop} at \uni{02C1} vs.\@ the <ˤ> \textsc{modifier
       letter small reversed glottal stop} at \uni{02E4}. Both are denoted in
       the Unicode Standard as the \textsc{pharyngealized diacritic} and both
       appear in various resources representing phonetic data online. This is
       thus an example for which the Unicode Standard does not solve the
       linguistic standardization problem. 
	\item There is at least one case in which the character name assigned by the
       Unicode Consortium does not match the IPA's description. In the Unicode
       Standard the <ǃ> at \uni{01C3} is labeled \textsc{latin letter retroflex
       click}, but in IPA is an alveolar or postalveolar click (not retroflex).
       This naming is probably best seen as a simple error in the Unicode
       Standard. This character is of course often simply typed as <!>
       \textsc{exclamation mark} at \uni{0021}.
\end{itemize}

@@

It is not just the Unicode Standard that offers multiple options for encoding 
the IPA. Even the IPA itself is not consistent in how specific meaningful 
transcriptions have to be encoded. There are a few cases in which the IPA 
explicitly allows for different options of transcribing the same phonetic 
content

Another example
we commonly encounter is the use of <g> \textsc{latin small letter g} at \uni{0067},
instead of the Unicode Standard IPA character for the voiced velar
stop <ɡ> \textsc{latin small letter script g} at \uni{0261}. One begins to question
whether this issue is at all apparent to the working linguist, or if they simply
use the \uni{0067} because it is easily keyboarded and thus saves time, whereas
the latter must be cumbersomely inserted as a special symbol in most
software.

\begin{itemize}
  \item tone marks (prefer tone symbols)
  \item marking of <g>\footnote{This issue was recently addressed by the International
Phonetic Association, which has taken the stance that both the keyboard
\textsc{latin small letter g} and the \textsc{latin small letter script g} are
valid input characters for the voiced velar plosive. Unfortunately, this
decision further introduces ambiguity for linguists trying to adhere to a strict
Unicode Standard IPA encoding.}
  \item marking of voiceless with ring diacritic
  \item tie bar
\end{itemize}
